---
title: "MB5370 Module 04. Workshop 1 - Introduction"
author: "Ben Cresswell"
output: html_document
date: "2024-03-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Install and load tidyverse packages
```{r}
## install.packages("tidyverse") # Delete this line once installed
library("tidyverse")
```


# Load the data (mpg is built into ggplot2)
```{r}
data(mpg)
```


# Quick data checks
```{r}
head(mpg)
glimpse(mpg)
summary(mpg)
```


# Create first plot
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

The plot shows a negative relationship between engine size (displ) and fuel efficiency (hwy). In other words, cars with big engines use more fuel. What does this say about fuel efficiency and engine size?



# Understanding grammar of graphics
What happens if we just try to run ggplot on its own?
```{r}
ggplot() 
# Or with just the data specified?
ggplot(data = mpg)
```

We need to map some aesthetics!!

# When you’re creating a plot, you essentially need two attributes of a plot: a geom and aesthetics.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```


# Change point colour by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```

# Change point size by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```
Note the warning!!!


# Change transparency (alpha) by class
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```
Another warning!!
Question: When would using alpha (or size be appropriate??)

# Change point shape by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```


# Make all points blue
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

# Troubleshooting 

# Faceting

## facet_wrap
Split the plot out by car type (class)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```


## facet_grid
A separate facet for each combination of drive-type (e.g. 4WD) * number of cylinders
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```
Note that there are no occurrences of 5 cylinder 4WDs OR RWD vehicles!


## facet_grid by just row (or column)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(cyl ~ .)
  #facet_grid(. ~ cyl) # Alternatively

```
#Facet_grid(. ~ cyl) Alternatively

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ cyl)

```
# Exercise:
Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? 

1) "nrow" dictates the number of rows in the layout of the individual panels, like "ncol" for the number of columns.
2) "dir" dictates the order of the individual panels (left --> right; top --> bottom etc)
3) "space" dictates whether panel sizes stay the same or stretch with the data ranges
4)"scales" dictates whether axes stay the same across panels
5) "drop" dictates if unused factor levels are hidden in the final graph or allowed to show
6) "axes" dictates whether axes are drawn only on the outside or also inside the grid
7) "axis.labels" decides where tick labels appear (only on the outside, or everywhere)
8) "labeller" dictates how the facet labels are formatted and styled
9) "switch" decides if labels appear at the bottom or left instead of the default spots
10) "shrink" decides whether panels zoom in to the summary data or keep the full range.
11) "facets" dictates which variable(s) are used to split the data into panels.

# Lines
We already used points:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```


However, ggplot2 can use a variety of geom objects to represent the data. Here, we might want to use bar plots, line charts, boxplots and so on. Well we can handle this issue in ggplot directly using a different geom to plot the same data. Here, instead of plotting points, we will use a smooth line. 

# To display the same data as a smooth line fit through the points use geom_smooth().
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```
### Question: when don't we use geom_line() here? What would that look like? ###

RECAP: A geom is an object that your plot uses to represent the data. To change the geom type in your plot, simply change the geom function that you add to your plot template. Sometimes you may want to try a few things out, in which case you could use comments to help you remember what worked and what didn’t. 

## Using comments (#)
```{r}
ggplot(data = mpg) + 
  #geom_point(mapping = aes(x = displ, y = hwy)) + # points horrible 
  geom_smooth(mapping = aes(x = displ, y = hwy)) # try smooth line
```

Question: how does R work out where to plot the line??? Can use the chunk output to inform us. Can also use the help menu.

# Changing linetype ( use a variable to control it)
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv)) # Can also use "lty = " for the same result
```


# Grouping objects by a categorical variable to draw multiple objects
```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```
 
# Change line colour based on drv value
```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE,
  )
```

# Multiple geoms
We already did this one way
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

A better way...
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```


# Can still manipulate each geom/layer separately:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```

Now we have different colours to indicate the raw data (geom_point) and one colour for the overall relationship (geom_smooth). Neat!

# The same goes if you want to specify different data for each layer. 
Use a filter (class = "subcompact") to select a subset of the data and plot only that subset. 
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```
Question: what does the "se = FALSE" do? It tells geom_smooth() not to draw the 95% confidence band around the smooth line.

# Exercise:
1.  What geom would you use to draw a line chart? geom_line(). A boxplot? geom_boxplot(). A histogram? geom_histogram(). An area chart? geom_area().
2.  Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_line()

ggplot(data = mpg, mapping = aes(x = class, y = hwy, fill = class)) +
  geom_boxplot(alpha = 0.7) +
  geom_point(position = position_jitter(width = 0.15), colour = "black", size = 1.1) +
  guides(fill = "none")

ggplot(data = mpg, mapping = aes(x = displ)) +
  geom_histogram()

ggplot(data = mpg, mapping = aes(x = displ)) +
  geom_histogram() +
  facet_wrap(~ class, ncol = 3)

ggplot(data = mpg, mapping = aes(x = displ, fill = class)) +
  geom_area(stat = "bin", position = "stack", bins = 20, alpha = 0.7)

```


3.  Will these two graphs look different? Why/why not?
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```
The two code blocks produce identical looking figures, but the difference lays in the fact that in the first plot there is aes(x= displ, y= hwy) set in mapping globally then adding geom_point() and geom_smooth(), while in the second the same data/mappings are given per each layer specifically.

# Transformations and stats

## Load data
```{r}
data("diamonds")
glimpse(diamonds)
```

# Bar plot
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```
On the x-axis we have the "cut" variable from the "diamonds" data set, whileon the y-axis the bin counts for each type of cut.

# Overriding defaults
Make some new data
```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo
```


```{r}
ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```
Question: Why did we specify now specify a y axis? We stored counts in "freq" data set, so compared to the previous bar plot where we were plotting raw data, now were have to specify y = freq in order to use the numbers instead of counting rows.

# What if we want to know the proportion rather than the absolute frequency?
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))
```
In this bar plot we are displaying the proportion of the total diamond data set rather that a count.

# Plotting statistical details
```{r}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```


# Aesthetic adjustments adjustments
Another way to boost the way you can convey information with plots using ggplot2 is to use aesthetics like colour or fill to change aspects of bar colours. We already did this once, but there are multiple options available to you, including changing the fill our outline colours.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```
Question: Does anyone notice anything different in the colour scheme of this plot?
(Hint: It's in the viridis colour palette (colour blind friendly), but why is it different from the colour palette we used earlier?) 

It looks different because in the first bar plot we changed colors for the outline of the bars, while in the second bar plot we changes the colors of the inside fill, even though we use viridis palette in both.

Check out the difference:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) 

ggplot(data = diamonds, mapping = aes(x = color, y = depth)) + 
  geom_point(mapping = aes(color = cut)) 
```
In the first plot, different colors indicates vehicle class for each engine size (numeric) on the x-axis, so the points are spread out. In the second plot, different colors indicates the cut of the diamond, organizing the points in vertical lines according to the category on the x-axis.

# Filling by a variable
Now try using these aesthetics to colour by another variable like clarity. Notice how the stacking is done automatically. This is done behind the scenes with a position argument.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```


# To alter transparency (alpha)
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")
```


# To color the bar outlines with no fill color
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```


# Position adjustments
position = "fill" works like stacking, but makes each set of stacked bars the same height.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```


# position = "dodge" 
Places overlapping objects directly beside one another.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

# Jittering
position = "jitter" adds a small amount of random noise to each point to avoid overplotting when points overlap. This is useful for scatterplots but not barplots.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")

```
```{r}
setwd(('C:/Users/miria/OneDrive - James Cook University/Github/MB5370_Mod04_Personal/data'))

```
# ASSIGNMENT: PLOT DECONSTRUCTION

# 1) Original Plot
This chart shows [main relationship/message] for [population/region/time]. However, it [problem 1] and [problem 2], which can mislead the reader about [effect/context]. Data from [source + year(s)]; I digitised values where needed.

```{r}

library(tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)
library(readr)
library(gridExtra)

setwd(('C:/Users/miria/OneDrive - James Cook University/Github/MB5370_Mod04_Personal/data'))
faostat <- read_csv2("data/faostat_bananas.csv",   # or the right path
                     show_col_types = FALSE,
                     col_types = cols(
                       Area  = col_character(),
                       Year  = col_integer(),
                       Value = col_double()
                     ))

faostat_bananas <- readr::read_csv("../data/faostat_bananas.csv", show_col_types = FALSE)

bananas <- gather(faostat_bananas, "Area", "Year", "Value")

# basic cleaning
faostat <- ban |>
  mutate(
    Area = as.factor(Area),
    Year = as.integer(Year),
    Value = as.double(Value)
  )

# total exports by year (for % shares and reference line)
ban_tot <- ban |>
  group_by(Year) |>
  summarise(Total = sum(Value, na.rm = TRUE), .groups = "drop")
ban_tot
# pick "key" exporters (top 4 by mean exports across all years)
top_exporters <- ban |>
  group_by(Area) |>
  summarise(mean_t = mean(Value, na.rm = TRUE), .groups = "drop") |>
  slice_max(mean_t, n = 4) |>
  pull(Area)

ban <- ban |>
  left_join(ban_tot, by = "Year") |>
  mutate(share = Value / Total)

# dataframe with last year for direct endpoint labels
end_year <- max(ban$Year, na.rm = TRUE)
end_pts <- ban |>
  filter(Year == end_year) |>
  arrange(desc(Value))
```










